A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-19 tex/codatadp.tex
theory def, cleanup.

TODO: Andy, introduce theory here. There were some ideas below in the comments
you might want to mention either here or in Section 3. I'm not sure what's the
best way to justify it though, but for the conference at least we can just claim it. (we
won't have the place for a proof anyway).

@@

\begin{lemma}[Refutation Soundness]%
\label{lem:rs}%
\afterDot
If there exists a closed derivation tree with root node $\Ec$, then $\Ec$ is unsatisfiable in $\thD$.
\end{lemma}
\begin{proof}
We prove this by structural induction on the (closed) derivation tree with root node $\Ec$.
First, if the derivation tree is an application of \rn{Conflict}, \rn{Clash}, or \rn{Cyclic},
then $\Ec$ is unsatisfiable in $\thD$.
For \rn{Conflict}, this is a consequence of equality reasoning.
For \rn{Clash}, this is a consequence of distinctness.
For \rn{Cyclic}, our construction of $\Val$ indicates that the class of values that $t_1$ can take in models of $\Ec$ is infeasible,
and thus $\Ec$ is unsatisfiable.
If the children of $\Ec$ are closed derivation trees 
whose root node is the result of applying the rule \rn{Split} on term $t$ of type $\tau$,
then by the induction hypothesis $\Ec \cup t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$ is unsatisfiable
for each $C_i \in \Ctr^\tau$.
Since by exhaustiveness, all models of $\thD$ entail (exactly) one equality $t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$,
we know $\Ec$ is unsatisfiable in $\thD$.
Otherwise, the child of $\Ec$ is a closed derivation tree 
whose root node is $\Ec \cup t_1 \teq t_2$ obtained by applying one of the rules \rn{Refl}, \rn{Sym}, \rn{Trans}, \rn{Cong}, \rn{Decomp}, \rn{Bisimilar}, or \rn{Singleton}.
In all cases, we have that $\Ec \models_{\thD} t_1 \teq t_2$.
For \rn{Refl}, \rn{Sym}, \rn{Trans}, \rn{Cong}, this is a consequence of equality reasoning.
For \rn{Decomp}, this is a consequence of injectivity.
For \rn{Bisimilar}, our construction of $\Val$ indicates that the value of $t_1$ and $t_2$ are equivalent in all models of $\Ec$.
For \rn{Singleton}, clearly $t_1$ and $t_2$ must have the same value since the cardinality of their type is one.
By the induction hypothesis, we have $\Ec \cup t_1 \teq t_2$ is unsatisfiable in $\thD$, 
and thus $\Ec$ is unsatisfiable in $\thD$.
\qed
\end{proof}

@@

First, we need the following definition.
We say that $t \vsimv{x} s$ if and only if $\mu$-terms $t$ and $s$ are syntactically equivalent
for some renaming that avoids capturing any variable other than $x$.
For instance, 
$\mu x. \const{D}( x ) \vsimv{y} \mu x. \const{D}( y )$,
$\mu x. \const{C}( x, x ) \vsimv{x} \mu y. \const{C}( x, y )$ and
$\mu x. \const{C}( z, x ) \vsimv{z} \mu y. \const{C}( z, y )$,
but
$\mu x. \const{D}( x ) \not\vsimv{x} \mu x. \const{D}( y )$ and
$\mu x. \const{C}( x, x ) \not\vsimv{y} \mu y. \const{C}( x, y )$.
For a variable $x$ of type $\tau$ and a fixed normal model $\M$,
we write $\mathcal{T}^\M_x( \Val )$ to denote the set consisting of all values $v$ from $\M( \tau )$
such that $v \vsimv{x} t_1\interp{t_2}$ for some term $t_1$ with subterm $t_2$ occurring in the range of $\Val$.
We first construct a mapping $\Val'$ by exhaustively applying the following update to $\Val$:

\(
\inferrule{
  x : \tau \in FV( \Val ) 
  \quad
  (\mu x. t) \aequiv v
  \quad
  v \in \M( \tau )
  \quad
  v \not\in \mathcal{T}^\M_x(\Val)
}{
  \Val := \Val \{ x \mapsto \mu x. t \}
}
\)

In this update, we choose an unassigned variable in $\Val$
and assign it a value that is not $\alpha$-equivalent to one occurring in $\mathcal{T}^\M_x(\Val)$.
Since this update removes one variable from the set $FV( \Val )$ and does not add any additional variables to $FV( \Val )$, 
it can only be applied a finite number of times.
We then let $\Val^\ast \ec{x}$ be $\nf{(\Val' \ec{x})}$ for all $\ec{x}$.

@@

To construct a completion of $\Val$, we must choose values for the free variables of $\Val$, $v_{\ec{x}}$ and $v_{\ec{w_3}}$.
The content of the set $\mathcal{T}^\M_{v_{\ec{x}}}( \Val )$ is $\alpha$-equivalent to:
\[%\begin{array}{c}
\{
\mu x. \const{A}( \const{B}( x )),
\const{A}( \mu y. \const{B}( y )),
\mu x. \const{B}( x ),
\mu x. \const{C}( \const{A}( x )),
\mu x. \const{A}( \const{C}( x )),
\mu x. \const{A}( x )
\}
%\end{array}
\]
Now, consider a model $\M$ that interprets variables in $\Ec$ based on the mapping $\Val$,
for instance, $x$ is interpretted as $\Val \ec{x}$, $y_1$ as $\Val \ec{y_1}$ and so on.
Assigning a value for $v_{\ec{x}}$ that is $\alpha$-equivalent to a value in $\mathcal{T}^\M_{v_{\ec{x}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent, 
which in turn may cause $\M$ to be inconsistent.
For instance, say we assign $\mu v_{\ec{x}}. \const{B}( v_{\ec{x}} )$ for $v_{\ec{x}}$.
After this substitution, $\Val \ec{y_2}$ is $\mu v_{\ec{y_2}}. \const{B}( \mu v_{\ec{x}}. \const{B}( v_{\ec{x}} ) )$, 
which has normal form $\mu v_{\ec{y_2}}. \const{B}( v_{\ec{y_2}} )$,
which is $\alpha$-equivalent to $\mu v_{\ec{x}}. \const{B}( v_{\ec{x}} )$.
However, this contradicts the disequality $x \not\teq y_2$ in $\Ec$.
Inversely, if the value assigned to $v_{\ec{x}}$ is not $\alpha$-equivalent to any term in $\mathcal{T}^\M_{v_{\ec{x}}}( \Val )$,
then the values in the range of $\Val$ remain $\alpha$-disequivalent.
For instance, we may assign values such as 
$\mu v_{\ec{x}}. \const{C}( v_{\ec{x}} )$,
$\mu v_{\ec{x}}. \const{C}( \mu x_1. \const{A}( x_1) )$, or
$\mu v_{\ec{x}}. \const{B}( \const{A}( v_{\ec{x}} ) )$
to $v_{\ec{x}}$.
Notice that legal substitutions for $v_{\ec{x}}$ may cause the range of $\Val$ to contain non-normal terms,
as in after applying the third substitution mentioned above,
$\Val \ec{w_1}$ is $\mu v_{\ec{w_1}}. \const{A}( \mu v_{\ec{x}}. \const{B}( \const{A}( v_{\ec{x}} ) ) )$.
$\Box$
\end{example}

@@

\begin{lemma} 
\label{lem:model-completion}
If $\Val$ is constructed for a saturated $\Ec$,
and $\Val^\ast$ is a completion of $\Val$ for normal model $\M$, then:
\begin{enumerate}
\item $\Val^\ast \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$, for all $\ec{x}$ of type $\tau$ of $\Ec$,
\item
If $\const{C}( t_1, \ldots, t_n ) \in \ec{x}$,
then $\Val^\ast \ec{x} = \mu v_{\ec{x}}. \const{C}( w_1, \ldots, w_n )$
and $\Val^\ast \ec{x} \interp{ w_i } = \Val^\ast \ec{t_i}$ for $i = 1, \ldots, n$.
%and $\muvar( w_i ) = v_{\ec{t_i}}$ for $i = 1, \ldots, n$,
%\item
%$t_1 \interp{ t_2 } = \Val^\ast \ec{x}$
%for all $t_1$ with subterm $t_2$ in the range of $\Val^\ast$ where $\muvar(t_2)=v_{\ec{x}}$,
\item 
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we first show that $\Val^\ast$ contains no free variables.
Assume by contradiction that $\Val^\ast$ contained a free variable $v_{\ec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else $v_{\ec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be one.
Since \rn{Singleton} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}^\M_x(\Val^\ast)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to $v_{\ec{x}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction, 
since there are only a finite number of closed terms in $\mathcal{T}^\M_x(\Val^\ast)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_x(\Val^\ast)$ to $v_{\ec{y}}$.
By construction, $\Val^\ast \ec{x}$ is normal.
Since \rn{Cyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\ast$ applies substitutions of the form 
$\{ y \mapsto t \}$, where $t$ is acyclic when the type of $y$ is not a codatatype.
Thus, $\Val^\ast \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\ast \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$.

@@

\begin{lemma}[Solution Soundness]%
\label{lem:ss}%
\afterDot
If there exists a derivation tree with root node $\Ec$ containing a saturated node, then $\Ec$ is satisfiable in $\thD$.
\end{lemma}
\begin{proof}
Let $\Fc$ be a saturated node in a derivation tree with root node $\Ec$.
%We will construct a model $\M$ for a set of equalities $\Fc$ that is equivalent to $\Fc_0$, and where 
%all equivalence classes of $\Fc$ contain at least one variable, and all selectors in $\Fc$ are applied to variables only.
%The former comes with no loss of generality since new equalities of the form $y \teq t$ for fresh variable $y$ can be added to $\Fc$ without affecting its satisfiability,
%The latter also comes with no loss of generality since nested applications of selectors in $\Fc_0$ can be replaced by fresh variables while adding additional equalities to $\Fc$.
We consider a normal model $\M$ 
that interprets wrongly-applied selectors based on equality information in $\Fc$,
and interprets the variables of $\Fc$ based on the completion $\Val^\ast$ of the mapping $\Val$ from step two of our calculus.
For the latter, let $\M( x )$ be the value in $\M$ that is $\alpha$-equivalent with $\Val^\ast \ec{x}$ for each variable $x$ in $\mathcal{T}(\Fc)$.

We first show $\M$ satisfies all equalities in $t_1 \teq t_2 \in \Fc$, where $t_1$ and $t_2$ have type $\tau$.
We show,
by structural induction on $t$,
that $\M( t ) \aequiv \Val^\ast \ec{t}$ for all terms $t \in \mathcal{T}( \Fc )$,
which since $\M$ is normal implies $\M \models t_1 \teq t_2$.

@@

We now show that all disequalities in $\Fc$ are satisfied by $\M$.
Say $t_1 \tneq t_2 \in \Fc$.
Since \rn{Conflict} does not apply, $t_1 \teq t_2 \not\in \Fc$ and thus $\ec{t_1}$ and $\ec{t_2}$ are distinct.
Since $\M( t_1 ) \vsim \Val^\ast \ec{t_1}$, $\M( t_2 ) \vsim \Val^\ast \ec{t_2}$
and Lemma~\ref{lem:model-completion}.3, $\M( t_1 ) \neq \M( t_2 )$, and thus $\M \models t_1 \tneq t_2$.

@@

\paragraph{Integration into DPLL(T)}
The calculus is implemented as \emph{theory solver} of CVC4,
that is, a specialized solver for determining the satisfiability of conjunctions of constraints for its theory.
Given a theory $T = T_1 \cup \ldots T_n$ and a input set of clauses $F$ in CNF,
the DPLL(T) \rem{CDCL(T)?} procedure
(incrementally) builds partial assignments from the atomic formulas of $F$ to truth values such that no clause in $F$ is falsified.
We may interpret this partial assignment as a set of literals where $a \in M$ for all atoms $a$ that it assigned the value $\top$, and $\neg a \in M$ for all atoms that it assigns the value $\bot$.
Then, at a high level, by the Nelson-Oppen combination framework,
each $T_i$-solver for $i = 1, \ldots, n$ takes as input a combination of (1) the purified form of $T_i$-constraints occurring in $M$ where terms containing symbols not belonging to $T_i$ are replaced by fresh variables of the same type, 
(2) additional equalities and disequalities between variables of types not belonging to $T_i$.
Let us call this set $M_i$.
Each $T_i$ solver either 
reports that a subset $C$ of $M_i$ is $T_i$-unsatisfiable in which case $\neg C$ is added as a additional clause to $F$,
adds an additional clause $C$ to $F$,
or does nothing.
When $M$ is a complete assignment for $F$, then a theory solver can only choose to do nothing when $M_i$ is indeed $T_i$-satisfiable.

Assume $\Ec$ is initially the set $M_i$ described above.
For each equality added to $\Ec$, we associate a set of equalities from $M_i$ that together $\thD$-entail $t \teq s$,
which we call its \emph{explanation}.
Similarly, each $\Val \ec{x}$ is also associated an explanation, that is, 
a set of equalities from $M_i$ that $\thD$-entail that the values of $\ec{x}$ in models of $\Ec$ are of the form $\Val \ec{x}$.
For instance, if $x \teq \const{C}( x ) \in M_i$, then $x \teq \const{C}( x )$ is a (possible) explanation for $\Val \ec{x} = \mu v_{\ec{x}}. \const{C}( v_{\ec{x}} )$.
Thus, the rules of the calculus are implemented as follows.
For all rules with conclusion $\bot$,
we report the union of the explanations for all premises is $\thD$-unsatisfiable.
For the \rn{Split} rule, we add a clause of the form
$(\const{C}_1( s^1_1( t ), \ldots, s^{a_1}_1( t ) ) \vee \ldots \vee t \teq \const{C}_n( s^1_n( t ), \ldots, s^{a_n}_n( t ) ))$
to $F$.
Notice that decisions on which branch to take are thus performed externally by the SAT solver, and not by the theory solver.
All other rules add equalities to the internal state of the theory solver.
The rules in step 1 of the calculus (along with the optimization for collapsing selectors) and the \rn{Singleton} rule are performed eagerly, 
that is for partial satisfying assignments $M$, while the rules in step 2 and the \rn{Split} rule are performed only for complete satisfying assignments $M$, 
and in that order.

>>>>

%TODO: Andy, introduce theory here. There were some ideas below in the comments
%you might want to mention either here or in Section 3. I'm not sure what's the
%best way to justify it though, but for the conference at least we can just claim it. (we
%won't have the place for a proof anyway).

Given a signature $\Sig$, we write $\thD$ to refer to the \emph{theory} of (co-)datatypes,
which in addition to $\Sig$ defines a class of $\Sig$-interpretations $\mathbf{I}$,
namely the ones that satisfy the axioms mentioned in this section.
In particular, this means that each interpretation in $\mathbf{I}$ has a fixed interpretation
for constructor terms and correctly-applied selector terms, but may differ on its interpretation for 
wrongly-applied selector terms and variables.
A formula $\varphi$ is $\thD$-satisfiable if there is an interpretation in $\mathbf{I}$ that satisfies $\varphi$,
and $\thD$-unsatisfiable otherwise.
\rem{more?}

@@

\begin{lemma}[Refutation Soundness]%
\label{lem:rs}%
\afterDot
If there exists a closed derivation tree with root node $\Ec$, then $\Ec$ is $\thD$-unsatisfiable.
\end{lemma}
\begin{proof}
We prove this by structural induction on the (closed) derivation tree with root node $\Ec$.
First, if the derivation tree is an application of \rn{Conflict}, \rn{Clash}, or \rn{Cyclic},
then $\Ec$ is $\thD$-unsatisfiable.
For \rn{Conflict}, this is a consequence of equality reasoning.
For \rn{Clash}, this is a consequence of distinctness.
For \rn{Cyclic}, our construction of $\Val$ indicates that the class of values that $t_1$ can take in models of $\Ec$ is infeasible,
and thus $\Ec$ is unsatisfiable.
If the children of $\Ec$ are closed derivation trees 
whose root node is the result of applying the rule \rn{Split} on term $t$ of type $\tau$,
then by the induction hypothesis $\Ec \cup t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$ is unsatisfiable
for each $C_i \in \Ctr^\tau$.
Since by exhaustiveness, all models of $\thD$ entail (exactly) one equality $t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$,
we know $\Ec$ is $\thD$-unsatisfiable.
Otherwise, the child of $\Ec$ is a closed derivation tree 
whose root node is $\Ec \cup t_1 \teq t_2$ obtained by applying one of the rules \rn{Refl}, \rn{Sym}, \rn{Trans}, \rn{Cong}, \rn{Decomp}, \rn{Bisimilar}, or \rn{Singleton}.
In all cases, we have that $\Ec \models_{\thD} t_1 \teq t_2$.
For \rn{Refl}, \rn{Sym}, \rn{Trans}, \rn{Cong}, this is a consequence of equality reasoning.
For \rn{Decomp}, this is a consequence of injectivity.
For \rn{Bisimilar}, our construction of $\Val$ indicates that the value of $t_1$ and $t_2$ are equivalent in all models of $\Ec$.
For \rn{Singleton}, clearly $t_1$ and $t_2$ must have the same value since the cardinality of their type is one.
By the induction hypothesis, we have $\Ec \cup t_1 \teq t_2$ is $\thD$-unsatisfiable, 
and thus $\Ec$ is $\thD$-unsatisfiable.
\qed
\end{proof}

@@

First, we need the following definition.
We say that $t \vsimv{x} s$ if and only if $\mu$-terms $t$ and $s$ are syntactically equivalent
for some renaming that avoids capturing any variable other than $x$.
For instance, 
$\mu x. \const{D}( x ) \vsimv{y} \mu x. \const{D}( y )$,
$\mu x. \const{C}( x, x ) \vsimv{x} \mu y. \const{C}( x, y )$ and
$\mu x. \const{C}( z, x ) \vsimv{z} \mu y. \const{C}( z, y )$,
but
$\mu x. \const{D}( x ) \not\vsimv{x} \mu x. \const{D}( y )$ and
$\mu x. \const{C}( x, x ) \not\vsimv{y} \mu y. \const{C}( x, y )$.
For a variable $x$ of type $\tau$ and a fixed normal model $\M$,
we write $\mathcal{T}^\M_x( \Val )$ to denote the set consisting of all values $v$ from $\M( \tau )$
such that $v \vsimv{x} t_1\interp{t_2}$ for some term $t_1$ with subterm $t_2$ occurring in the range of $\Val$.
We construct $\Val^\ast$ by exhaustively applying the following update to $\Val$:

\(
\inferrule{
  x : \tau \in FV( \Val ) 
  \quad
  (\mu x. t) \aequiv v
  \quad
  v \in \M( \tau )
  \quad
  v \not\in \mathcal{T}^\M_x(\Val)
}{
  \Val := \nf{(\Val \{ x \mapsto \mu x. t \})}
}
\)

In this update, we choose an unassigned variable in $\Val$
and assign it a value that is not $\alpha$-equivalent to one occurring in $\mathcal{T}^\M_x(\Val)$.
Since this update removes one variable from the set $FV( \Val )$ and does not add any additional variables to $FV( \Val )$, 
it can only be applied a finite number of times.
After assigning this value, we normalize all terms in the range of $\Val$.
%We then let $\Val^\ast \ec{x}$ be $\nf{(\Val' \ec{x})}$ for all $\ec{x}$.

@@

To construct a completion of $\Val$, we must choose values for the free variables of $\Val$, $v_{\ec{x}}$ and $v_{\ec{w_3}}$.
The content of the set $\mathcal{T}^\M_{v_{\ec{x}}}( \Val )$ is $\alpha$-equivalent to:
\[%\begin{array}{c}
\{
\mu x. \const{A}( \const{B}( x )),
\const{A}( \mu y. \const{B}( y )),
\mu x. \const{B}( x ),
\mu x. \const{C}( \const{A}( x )),
\mu x. \const{A}( \const{C}( x )),
\mu x. \const{A}( x )
\}
%\end{array}
\]
Now, consider a model $\M$ that interprets variables in $\Ec$ based on the mapping $\Val$,
for instance, $x$ is interpretted as $\Val \ec{x}$, $y_1$ as $\Val \ec{y_1}$ and so on.
Assigning a value for $v_{\ec{x}}$ that is $\alpha$-equivalent to a value in $\mathcal{T}^\M_{v_{\ec{x}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent, 
which in turn may cause $\M$ to be inconsistent.
For instance, say we assign $\mu v_{\ec{x}}. \const{B}( v_{\ec{x}} )$ for $v_{\ec{x}}$.
After this substitution, $\Val \ec{y_2}$ is $\mu v_{\ec{y_2}}. \const{B}( \mu v_{\ec{x}}. \const{B}( v_{\ec{x}} ) )$, 
which has normal form $\mu v_{\ec{y_2}}. \const{B}( v_{\ec{y_2}} )$,
which is $\alpha$-equivalent to $\mu v_{\ec{x}}. \const{B}( v_{\ec{x}} )$.
However, this contradicts the disequality $x \not\teq y_2$ in $\Ec$.
Inversely, if the value assigned to $v_{\ec{x}}$ is not $\alpha$-equivalent to any term in $\mathcal{T}^\M_{v_{\ec{x}}}( \Val )$,
then the values in the range of $\Val$ remain $\alpha$-disequivalent.
For instance, we may assign values such as 
$\mu v_{\ec{x}}. \const{C}( v_{\ec{x}} )$,
$\mu v_{\ec{x}}. \const{C}( \mu x_1. \const{A}( x_1) )$, or
$\mu v_{\ec{x}}. \const{B}( \const{A}( v_{\ec{x}} ) )$
to $v_{\ec{x}}$.
Notice that legal substitutions for $v_{\ec{x}}$ may cause the range of $\Val$ to contain non-normal terms,
as in after applying the substitution for the third term mentioned above,
$\Val \ec{w_1}$ is $\mu v_{\ec{w_1}}. \const{A}( \mu v_{\ec{x}}. \const{B}( \const{A}( v_{\ec{x}} ) ) )$,
which has normal form
$\mu v_{\ec{w_1}}. \const{A}( \mu v_{\ec{x}}. \const{B}( v_{\ec{w_1}} ) )$.
$\Box$
\end{example}

@@

\begin{lemma} 
\label{lem:model-completion}
If $\Val$ is constructed for a saturated $\Ec$,
and $\Val^\ast$ is a completion of $\Val$ for normal model $\M$, then:
\begin{enumerate}
\item $\Val^\ast \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$, where $\ec{x}$ has type $\tau$,
\item
If $\const{C}( t_1, \ldots, t_n ) \in \ec{x}$,
then $\Val^\ast \ec{x} = \mu v_{\ec{x}}. \const{C}( w_1, \ldots, w_n )$
and $\Val^\ast \ec{x} \interp{ w_i } = \Val^\ast \ec{t_i}$ for $i = 1, \ldots, n$.
%and $\muvar( w_i ) = v_{\ec{t_i}}$ for $i = 1, \ldots, n$,
%\item
%$t_1 \interp{ t_2 } = \Val^\ast \ec{x}$
%for all $t_1$ with subterm $t_2$ in the range of $\Val^\ast$ where $\muvar(t_2)=v_{\ec{x}}$,
\item 
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we first show that $\Val^\ast$ contains no free variables.
Assume by contradiction that $\Val^\ast$ contained a free variable $v_{\ec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else $v_{\ec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be one.
Since \rn{Singleton} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}^\M_x(\Val^\ast)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to $v_{\ec{x}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction, 
since there are only a finite number of closed terms in $\mathcal{T}^\M_x(\Val^\ast)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_x(\Val^\ast)$ to $v_{\ec{y}}$.
By construction, $\Val^\ast \ec{x}$ is normal.
Since \rn{Cyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\ast$ applies substitutions of the form 
$\{ y \mapsto t \}$, where $t$ is acyclic when the type of $y$ is not a codatatype.
Thus, $\Val^\ast \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\ast \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$.

@@

\begin{lemma}[Solution Soundness]%
\label{lem:ss}%
\afterDot
If there exists a derivation tree with root node $\Ec$ containing a saturated node, then $\Ec$ is $\thD$-satisfiable.
\end{lemma}
\begin{proof}
Let $\Fc$ be a saturated node in a derivation tree with root node $\Ec$.
%We will construct a model $\M$ for a set of equalities $\Fc$ that is equivalent to $\Fc_0$, and where 
%all equivalence classes of $\Fc$ contain at least one variable, and all selectors in $\Fc$ are applied to variables only.
%The former comes with no loss of generality since new equalities of the form $y \teq t$ for fresh variable $y$ can be added to $\Fc$ without affecting its satisfiability,
%The latter also comes with no loss of generality since nested applications of selectors in $\Fc_0$ can be replaced by fresh variables while adding additional equalities to $\Fc$.
We consider a normal model $\M$ 
that interprets wrongly-applied selectors based on equality information in $\Fc$,
and interprets the variables of $\Fc$ based on the completion $\Val^\ast$ of the mapping $\Val$ from step two of our calculus.
For the latter, let $\M( x )$ be the value in $\M$ that is $\alpha$-equivalent with $\Val^\ast \ec{x}$ for each variable $x$ in $\mathcal{T}(\Fc)$.

We first show $\M$ satisfies all equalities in $t_1 \teq t_2 \in \Fc$, where $t_1$ and $t_2$ have type $\tau$.
We show,
by structural induction on $t$,
that $\M( t ) \aequiv \Val^\ast \ec{t}$ for all terms $t \in \mathcal{T}( \Fc )$,
which since $\M$ is normal, implies $\M \models t_1 \teq t_2$.

@@

We now show that all disequalities in $\Fc$ are satisfied by $\M$.
Say $t_1 \tneq t_2 \in \Fc$.
Since \rn{Conflict} does not apply, $t_1 \teq t_2 \not\in \Fc$ and thus $\ec{t_1}$ and $\ec{t_2}$ are distinct.
Since $\M( t_1 ) \vsim \Val^\ast \ec{t_1}$ and $\M( t_2 ) \vsim \Val^\ast \ec{t_2}$,
by Lemma~\ref{lem:model-completion}.3, $\M( t_1 ) \neq \M( t_2 )$, and thus $\M \models t_1 \tneq t_2$.

@@

\paragraph{Integration into DPLL(T)}
The calculus is implemented as \emph{theory solver} of CVC4,
that is, a specialized solver for determining the satisfiability of conjunctions of constraints for its theory.
Given a theory $T = T_1 \cup \ldots T_n$ and a input set of clauses $F$ in CNF,
the DPLL(T) \rem{CDCL(T)?} procedure
(incrementally) builds partial assignments from the atomic formulas of $F$ to truth values such that no clause in $F$ is falsified.
We may interpret this partial assignment as a set of literals where $a \in M$ for all atoms $a$ that it assigned the value $\top$, and $\neg a \in M$ for all atoms that it assigns the value $\bot$.
Then, at a high level, by the Nelson-Oppen combination framework,
each $T_i$-solver for $i = 1, \ldots, n$ takes as input a combination of (1) the purified form of $T_i$-constraints occurring in $M$ where terms containing symbols not belonging to $T_i$ are replaced by fresh variables of the same type, 
(2) additional equalities and disequalities between variables of types not belonging to $T_i$.
Let us call this set $M_i$.
Each $T_i$ solver either 
reports that a subset $C$ of $M_i$ is $T_i$-unsatisfiable, in which case $\neg C$ is added as a additional clause to $F$,
adds an additional clause $C$ to $F$,
or does nothing.
When $M$ is a complete assignment for $F$, then a theory solver can only choose to do nothing when $M_i$ is indeed $T_i$-satisfiable.

Assume $\Ec$ is initially the set $M_i$ described above.
For each equality added to $\Ec$, we associate a set of equalities from $M_i$ that together entail $t \teq s$,
which we call its \emph{explanation}.
Similarly, each $\Val \ec{x}$ is also associated an explanation, that is, 
a set of equalities from $M_i$ that entail that the values of $\ec{x}$ in models of $\Ec$ are of the form $\Val \ec{x}$.
For instance, if $x \teq \const{C}( x ) \in M_i$, then $x \teq \const{C}( x )$ is a (possible) explanation for $\Val \ec{x} = \mu v_{\ec{x}}. \const{C}( v_{\ec{x}} )$.
Thus, the rules of the calculus are implemented as follows.
For all rules with conclusion $\bot$,
we report the union of the explanations for all premises is $\thD$-unsatisfiable.
For the \rn{Split} rule, we add a clause of the form
$(\const{C}_1( s^1_1( t ), \ldots, s^{a_1}_1( t ) ) \vee \ldots \vee t \teq \const{C}_n( s^1_n( t ), \ldots, s^{a_n}_n( t ) ))$
to $F$.
Notice that decisions on which branch to take are thus performed externally by the SAT solver, and not by the theory solver.
All other rules add equalities to the internal state of the theory solver.
The rules in step 1 of the calculus (along with the optimization for collapsing selectors) and the \rn{Singleton} rule are performed eagerly, 
that is for partial satisfying assignments $M$, while the rules in step 2 and the \rn{Split} rule are performed only for complete satisfying assignments $M$, 
and in that order.
